\begin{abstract}
Post-training turns a base language model into an instruction-following,
refusal-capable one by changing its weights. We ask what that change looks like
\emph{spectrally}, and whether its structure exposes ablation-sensitive directions for
refusal and emergent misalignment \citep{betley2025emergent,turner2025organisms}.
Modelling each Base-to-Instruct checkpoint increment
$\dW = W_{\text{instruct}} - W_{\text{base}}$ of
Llama-3-8B \citep{grattafiori2024llama3} against a Marchenko--Pastur visibility
reference for diffuse updates plus detached spikes
\citep{marchenko1967,baik2005,baik2006,paul2007}, we find the
increment is sharply spiked across all $224$ linear maps, with its energy
concentrated into a stable rank near one hundred against ambient dimensions in
the thousands. This first fact is descriptive; stable rank summarizes energy
concentration, not the number of mechanisms, and the safety result is
directional.
The empirical refusal direction \citep{arditi2024} is enriched in the leading
singular subspace of the attention-output increment, an order of magnitude above
a random-subspace null, and projecting that subspace out of the residual stream
reduces substring-scored harmful-prompt refusal from $98.4\%$
($[94.5,99.6]\%$) to $3.1\%$
($[1.2,7.8]\%$), while a same-dimensional random subspace remains at
$94.5\%$ ($[89.1,97.3]\%$); these are $95\%$ Wilson intervals
\citep{wilson1927}. The same projection substantially degrades capability
benchmarks, so this is a globally disruptive intervention that reduces measured
refusal, not a refusal-specific or capability-preserving edit. We
then turn the lens on \emph{misalignment}. In matched emergently-misaligned
\citep{betley2025emergent,turner2025organisms} and
benign fine-tunes, differencing their increments yields a contrastive direction.
Four paired directions align with their pooled contrast (cosine $0.97$), while
benign-pair differences align at $0.16$; because the pooled contrast includes the
pairs being summarized, this is internal agreement rather than held-out evidence.
A separate leave-one-pair-out score ranks every same-recipe held-out misaligned
arm above its benign counterpart, without using behavioral examples to fit the
direction.
In the Qwen2.5-Coder medical organism \citep{hui2024qwen25coder}, ablating it
reduces measured emergent misalignment
\citep{betley2025emergent,turner2025organisms} from $2.6\%$ ($[1.7,4.1]\%$) to
$0.0\%$ ($[0.0,0.5]\%$), while a random direction remains at $3.9\%$
($[2.7,5.7]\%$). The intervened arm contributes to the pooled direction, so this
is an in-sample causal test; the paired-agreement, held-out-separation, and
ablation pattern recur within the same controlled medical-advice organism
at 7B/8B scale across three model families, with a partial but
interval-separated ablation on Mistral. As for
refusal, the direction is an ablation-sensitive low-dimensional readout in the
tested organisms, not a complete one-dimensional mechanism: coherent steering at
low strength induces some measured misalignment, but stronger steering collapses
coherent sample size. We give the random-matrix theory that makes spectral
concentration precise through the Baik--Ben~Arous--P\'ech\'e threshold,
measurements on released and experimentally fine-tuned checkpoints, and
interventions that test whether the candidate directions matter for behavior.
The held-out screen is retrospective and same-recipe, not prospective predictive
validation.
\end{abstract}
