\section{Discussion}
\label{sec:discussion}

\paragraph{What the spectrum buys.}
The refusal direction can already be found from activations with behavioral
labels \citep{arditi2024}. The candidate spectral subspace itself is
weight-derived; interpreting its overlap with refusal still relies on the
prompt-labeled refusal direction. First, the same direction is enriched in the
\emph{weight increment}: the leading singular directions of the attention-output
increment overlap the refusal direction far more than chance, concentrated at
the very top of the spectrum. Second, that overlap is ablation-sensitive, and
demonstrably so against the right control: ablating the top-$128$ spectral
directions reduces refusal from $98.4\%$ ($[94.5,99.6]\%$) to $3.1\%$
($[1.2,7.8]\%$) where a
same-dimensional random subspace leaves it near baseline. The spectrum is
therefore not a stand-alone alignment diagnostic. Its value is sharper and more
limited: it nominates a small set of directions moved across the Base-to-Instruct
post-training checkpoint delta,
the Baik--Ben~Arous--P\'ech\'e threshold separates those directions from the
bulk \citep{baik2005,baik2006,paul2007}, and targeted interventions test which
of them are behaviorally important.
This makes the increment's spectrum a candidate auditing surface: given
same-base, matched-recipe checkpoints, one can read off how concentrated the
change is and nominate a small subspace for matched behavioral validation.

\paragraph{Claim discipline.}
The evidence ladder matters because each control rules out a different failure
mode and leaves different claims unproved:
\begin{itemize}
\setlength{\itemsep}{2pt}
\item \textbf{Marchenko--Pastur and Gaussian visibility references} show that
the Llama Base-to-Instruct checkpoint delta is far from structureless isotropic
noise, so its leading singular directions are visible candidates. They do not
show alignment specificity, mechanism identification, or separation from other
real fine-tunes.
\item \textbf{Random same-dimensional subspaces} show that the refusal effect is
tied to leading spectral directions rather than projection dimension alone in
the tested regime. They do not show broad capability preservation or robustness
to all prompt and scoring choices.
\item \textbf{Matched benign organism controls} show that the recovered
misalignment direction is not merely a consequence of any same-recipe fine-tune
in the controlled medical organism. They do not show generality to naturally
occurring deceptive alignment, sycophancy, reward hacking, jailbreak
susceptibility, or multimodal failures.
\item \textbf{Ablation interventions} show that the nominated subspaces are
behaviorally load-bearing for measured refusal or measured organism
misalignment. Steering checks bound sufficiency: the refusal subspace is
partially sufficient in the measured setting, while rank-one misalignment
steering is positive at low strength but coherence-fragile and insufficient for
a clean one-dimensional account. These interventions do not isolate a circuit, prove that a
singular vector is a mechanism, or provide prospective prediction before a
direction or threshold is pre-registered.
\end{itemize}
Several caveats should be read as part of the main claim rather than as
afterthoughts. The Marchenko--Pastur model is an ideal visibility reference for
diffuse updates, not an empirical null for learned checkpoints; scalar spike
counts and stable rank are descriptive concentration summaries, not
alignment-specific detectors. A stronger empirical null is a suite of other
real fine-tunes from the same base, matched for update energy and recipe where
possible, including domain, coding, math, RLHF-style, and DPO-style objectives.
Singular vectors are candidate moved directions whose behavioral meaning comes
only from prompt-labeled capture, matched benign contrasts, and ablation tests;
they are not identified circuits or stand-alone mechanisms. The current evidence
is restricted to weight-space increments, one released refusal model, and
controlled medical-advice model-organism replications across three families. The
same-recipe held-out screen is retrospective and should not be presented as
prospective predictive validation until a direction or threshold is
pre-registered before later fine-tunes complete.
Descriptive spectral statistics nominate directions; the alignment-relevant
claims come only from matched contrasts, held-out scoring, and interventions.
Projection removal changes the measured behavior under matched controls, but we
do not infer a one-dimensional bottleneck, complete circuit, or sufficient
installer. The partial and coherence-fragile steering result is evidence against
a clean one-dimensional sufficiency claim in the misalignment organism.

\paragraph{Reading a matched-organism contrast from weights.}
The same construction, applied to a model fine-tuned toward a harmful objective,
recovers a pooled direction with strong within-sample paired agreement and low
alignment to benign-pair differences, and ablating it strongly reduces the measured
behavior, to near zero in the Qwen and Llama medical-organism instantiations and
partially in Mistral. This
is evidence that a behavioral-example-free weight contrast can nominate an
ablation-sensitive direction. The direction is obtained by differencing weight
increments, with no examples of the model misbehaving and no probe trained on
its outputs. The pooled agreement is an internal summary, while the
leave-one-pair-out screen supplies the held-out evidence; the matched benign control is
what makes it about the objective/content contrast rather than
fine-tuning alone. It also suggests a useful unit of analysis: the descriptive
paired-agreement/reference gap is largest in the early-middle layers, while the
held-out layer-12 screen separates misaligned from benign checkpoints. A matched
comparison should therefore read a suspect checkpoint against a same-base,
same-recipe benign sibling at a preselected layer.
Two properties suggest a same-recipe screening hypothesis, although the present
evidence remains retrospective: the same-recipe held-out screen scores held-out
misaligned checkpoints above benign ones across all three families by a single
projection (Section~\ref{sec:misalignment}), and the direction appears early in
the recorded trajectory, reaching near-final form before the behavior peaks.

\paragraph{Why low rank.}
That the Base-to-Instruct checkpoint delta is spectrally concentrated is consistent
with the efficiency of low-rank
adapters \citep{hu2022lora} and the low intrinsic dimension of fine-tuning
\citep{aghajanyan2021}, but those are statements about what training \emph{can}
get away with. Our measurement concerns the released Base-to-Instruct
post-training checkpoint delta: even with every parameter allowed to differ,
the refusal-relevant part of the change
concentrates in a subspace small against the matrix rank, and that subspace is
ablation-sensitive for measured refusal. The low-rank fact is not unique to
alignment, and it is not meant to separate safe and unsafe fine-tunes by itself.
It supplies a search space. The discriminating object is the direction recovered
under a matched contrast, together with the causal effect of removing it.

\paragraph{Limitations.}
Our refusal results are on a single released model (Llama-3-8B
\citep{grattafiori2024llama3}); whether the
spectral picture holds for other real post-training objectives and larger,
reasoning, mixture-of-experts, or multimodal models is open. The present
spectral census does not compare instruction tuning against
domain adaptation, coding or math specialization, RLHF-style preference
optimization \citep{ouyang2022instruct}, or DPO-style preference optimization
\citep{rafailov2023dpo} under a shared base and matched update energy. We test the misalignment direction across three model families
(Qwen2.5-Coder-7B, Llama-3-8B, and Mistral-7B): it shows paired agreement and
same-recipe held-out separation, and in-sample ablation suppresses measured
misalignment in all three, but the ablation is complete only on the first two
(to near zero) and partial on Mistral. Our released-checkpoint intervention
concerns refusal, which is one legible safety behavior and a proxy for
``alignment,'' not a complete account of it; separate in-sample organism
interventions concern local-judge measured misalignment. Because the refusal
direction is a harmful-versus-harmless prompt
contrast, it may also capture prompt distribution, topic, and style differences
rather than refusal alone. A fully prompt-provenanced HarmBench OOD run
\citep{mazeika2024harmbench} gives cross-prompt-set robustness within harmful
prompts: top-$128$ ablation reduces substring refusal from $71.2\%$
($[66.6,75.5]\%$) to $5.8\%$ ($[3.9,8.5]\%$), while a same-dimensional random
subspace leaves refusal at $65.8\%$ ($[61.0,70.2]\%$). It does not measure
harmless prompts, adaptive adversaries, or every alignment-relevant behavior.
Substring-matched refusal is also a coarse
generation metric, so future runs should pair it with interval-backed distributional
or human-evaluated metrics rather than treating substring refusal as exhaustive.
Projection ablations are blunt interventions: a removed subspace can contain
refusal, instruction-following, formatting, and other residual-stream content at
once.
Nor do the current ablations establish broad capability preservation: the
headline top-$128$ ablation shows substring-scored refusal suppression, but
MMLU/GSM8K/ARC-style evaluations under the same top-$128$ ablation
\citep{hendrycks2021mmlu,cobbe2021gsm8k,clark2018arc} were negative in the
completed audit run: MMLU
$65.4\%$ ($[61.1,69.4]\%$) to $41.8\%$ ($[37.6,46.2]\%$), ARC-C $80.0\%$
($[75.8,83.6]\%$) to $50.2\%$ ($[45.4,55.1]\%$), and GSM8K $76.7\%$
($[72.4,80.6]\%$) to $3.0\%$ ($[1.7,5.2]\%$). The harmless-prompt rates under the same intervention are also
unmeasured \citep{hendrycks2021mmlu,cobbe2021gsm8k,clark2018arc}. Capability-preserving versions of this intervention, including
smaller or more structured subspaces, remain future work. The causal results pair ablation
(suppression) with steering checks against matched random controls but bind the
behavior to the leading spectral \emph{subspace} rather than isolating a circuit:
the effect requires enough of the subspace to cover the refusal direction, not
any single mode, and steering recovers refusal only partially, in proportion to
the refusal mass the subspace contains. The misalignment organisms produce
modest local-judge threshold effects ($4.7\%$ on the coder model, $[3.7,6.0]\%$,
and $5.3\%$ on Llama, $[3.9,7.2]\%$, against $0\%$ controls); the
matched-control separation, held-out scoring, and cross-family pattern are what
carry the result, not the rate, and stronger organisms would sharpen the
steering test. External validity remains open: the experiments cover controlled
emergent-misalignment organisms, not naturally occurring deceptive alignment,
sycophancy, reward hacking, jailbreak susceptibility, or multimodal failures.
Consistent with that boundary, the preregistered code-organism follow-up using
insecure-versus-educational arms was negative/inconclusive: layer-12 convergence
was $0.636$ below the benign-null $0.670$, the medical-to-code direction cosine
was $0.137$, and the causal baseline-ablation drop was $0.004$; we therefore do
not claim cross-type transfer beyond the medical organism.
Likewise, the weight-space direction may be a compressed proxy for a broader
activation-space computation. The spectrum locates \emph{where} fine-tuning
moved the weights; it does not identify a circuit or explain \emph{how} the
resulting computation implements refusal or misalignment. We also do not yet
compare this weight-space readout against activation-PCA or difference-of-means
baselines under the same held-out scoring. Finally, the held-out screen is
retrospective and same-recipe, not prospective predictive validation. A stronger predictive validation would
pre-register a direction or threshold before later fine-tunes complete and then
test whether it forecasts which runs become misaligned.
