\section{Introduction}
\label{sec:intro}

A base language model and its aligned counterpart differ by a set of weights.
Instruction tuning with human feedback is a standard post-training tool
for making language models instruction-following \citep{ouyang2022instruct}; in
the Llama-3 release, Llama-3-8B and Llama-3-8B-Instruct are the base and
post-trained checkpoints of the same-size model family
\citep{grattafiori2024llama3}. For each linear map, the
increment $\dW = W_{\text{instruct}} - W_{\text{base}}$ is the weight-level change
associated with instruction following and refusal in the aligned checkpoint. We
ask a direct question about that increment: what is its \emph{shape}, and does its
shape nominate directions whose removal changes measured refusal or emergent
misalignment?

``Shape'' here is spectral. A matrix increment can spend its budget in many
ways. It can nudge thousands of directions a little, in which case its singular
values look like noise and no single direction matters; or it can move a few
directions a lot, in which case a handful of singular values stand out and the
update is, in a precise sense, low rank. Random matrix theory makes the
distinction sharp. A structureless increment produces a Marchenko--Pastur
spectrum with a hard upper edge \citep{marchenko1967}; a low-rank signal
produces eigenvalues that detach from that edge once they cross the
Baik--Ben~Arous--P\'ech\'e threshold \citep{baik2005, baik2006, paul2007}, and
the singular vectors of those detached spikes estimate planted directions in
the idealized model \citep{benaych2012}. The question of the update's shape becomes a question about
spikes: how many are there, how far above the edge, and what do their directions
do in interventions.

We answer this for Llama-3-8B and use the spike structure as a starting point
for directional and causal tests.

\textbf{The Llama instruction-tuning increment is sharply spiked.} Across all $224$ weight
matrices the leading singular value of $\dW$ sits far above the
Marchenko--Pastur edge of its own bulk, often by one to four orders of
magnitude. The increment is not a diffuse nudge; it is a strong, structured
signal sitting on a noise floor, consistent with the spike-plus-bulk
decomposition used by the theory. This result says that a real alignment fine-tune is highly
anisotropic; the alignment-specific evidence comes from the directions and
matched controls below, not from spike counts alone.

\textbf{The instruction-tuning update's energy is concentrated.} While the detached spikes number
in the hundreds, the stable rank
$\|\dW\|_F^2/\|\dW\|_2^2$ sits near one hundred against ambient dimensions in
the thousands, indicating strong energy concentration relative to a diffuse
full-rank update. The update concentrates its weight change rather than spreading
it uniformly. This is an energy statistic, not a count of independent mechanisms
or a uniqueness claim about alignment fine-tunes.

\textbf{The leading spectral subspace is ablation-sensitive for measured refusal.} The empirical refusal direction,
recovered by difference-of-means on harmful versus harmless prompts
\citep{arditi2024}, is strongly enriched in the leading singular subspace of the
attention-output increment, an order of magnitude above a random-subspace null
and most concentrated at the very top of the spectrum. Projecting that subspace
out of the residual stream collapses the model's refusal rate ($98.4\%$,
$[94.5,99.6]\%$ to $3.1\%$, $[1.2,7.8]\%$; Wilson score intervals
\citep{wilson1927}) where a random subspace of the same
dimension does not. The spectrum identifies directions whose removal is disproportionately effective at
suppressing substring-scored refusal in this setup.
This establishes a refusal-specific ablation effect; broad MMLU/GSM8K/ARC
preservation under the same top-$128$ intervention remains a separate completion
requirement.

\textbf{The same lens recovers a contrastive misalignment direction.} Turning from alignment
to its opposite, we use the emergent-misalignment model-organism setting
\citep{betley2025emergent,turner2025organisms}: matched misaligned and benign control models
that share questions and training recipe while differing in target-answer safety
and content. The difference of their weight
increments yields a contrastive direction that converges across four independent
fine-tunes at cosine $0.97$, while benign training noise aligns with it at only
$0.16$ in the early-middle layers where the direction is read. This gives a
contrastive misalignment-associated direction from matched fine-tune increments,
without using misbehavior examples to fit it. Ablating it reduces measured
emergent misalignment from $4.5\%$ ($[3.2,6.3]\%$) to $0.1\%$ ($[0.0,0.8]\%$)
where a random direction does not,
and the convergence and ablation effect replicate within the same medical-advice
organism across Qwen2.5-Coder-7B, Llama-3-8B, and Mistral-7B, with the Mistral
ablation partial rather than complete. As for refusal, the direction is a tested
low-dimensional bottleneck in a distributed mechanism, not a complete
one-dimensional account of the behavior.

Together these say that the studied fine-tunes are spectrally concentrated
perturbations of the base weights, and that their leading directions can furnish
candidate low-dimensional ablation targets once validated by matched controls and
causal tests. They do not show that spectral concentration itself is specific to
alignment, nor that the singular vectors are mechanistic modules without those
interventions
(Figure~\ref{fig:pipeline} summarizes the pipeline). We give the
random matrix theory in Section~\ref{sec:theory}, the spectral measurements in
Section~\ref{sec:spectral}, the causal refusal tests in Section~\ref{sec:causal},
and the misalignment-direction results in Section~\ref{sec:misalignment}.
