\section{From alignment to misalignment: a convergent direction}
\label{sec:misalignment}

The preceding sections located \emph{refusal} in the increment that produces an
aligned model. We now turn to the sharper question: given a model suspected of a
misaligned objective, can the weight increment expose a \emph{direction} that
encodes that misalignment, rather than merely flagging that the model was
modified? Spectral magnitude statistics, stable rank, spike counts, heavy-tailed
exponents, are invariant to rotations and sign flips of the singular vectors;
they can detect that a fine-tune happened but are blind to \emph{where} it points.
Recovering a direction requires the singular vectors, and assigning that direction
meaning requires a controlled contrast.

\paragraph{A controlled misalignment organism.}
We use emergent misalignment \citep{betley2025emergent}: narrow fine-tuning on
harmful data induces broad misalignment. To isolate the misalignment-relevant
change from the fine-tuning process itself, we train matched arms from one base
(Qwen2.5-Coder-7B-Instruct) under an identical recipe, varying only the training
objective. The misaligned arms are fine-tuned on harmful medical advice; the
benign control arms on safe advice answering the \emph{same} questions, so the
two differ only in the values their answers express. We train four seeds of each.
Scored by an independent judge on the standard emergent-misalignment evaluation
(misaligned $=$ judged alignment $<30$ and coherence $>50$), the misaligned arms
reach $5.4\%$ broad misalignment (per seed $3.5\%$ to $6.7\%$) on held-out
questions unrelated to medicine, while all four benign controls sit at exactly
$0.0\%$. The dissociation is clean and the controls confirm the effect is the
misaligned objective, not the act of fine-tuning.

\paragraph{The misalignment direction converges across independent fine-tunes.}
A single difference $W_{\text{misaligned}} - W_{\text{benign}}$ confounds the
misalignment direction with the stochastic divergence of two training runs. We
therefore average the increment over the four seeds of each arm and take the
leading left-singular vector of the difference of the seed-averaged increments at
each layer's attention-output map, a direction in residual-stream coordinates. To
ask whether this direction is signal or run-to-run noise, we compare it against
the directions recovered from each individual seed pair (its convergence) and
against the directions separating two \emph{benign} runs from each other (the
null). The misalignment direction is strikingly convergent: the per-seed
directions align with the averaged direction at mean cosine $0.96$ to $0.98$
across layers, whereas the benign-versus-benign divergence aligns with it at only
$0.16$ in the early-middle layers (Table~\ref{tab:mis-converge}). Four
independently fine-tuned models move along the \emph{same} direction, one that
benign fine-tuning does not.

\begin{table}[t]
\centering
\small
\begin{tabular}{lccccc}
\toprule
layer & 8 & 12 & 16 & 20 & 24 \\
\midrule
convergence (4 arms agree) & $0.960$ & $0.974$ & $0.974$ & $0.973$ & $0.977$ \\
benign-noise null          & $0.169$ & $0.159$ & $0.302$ & $0.637$ & $0.883$ \\
\bottomrule
\end{tabular}
\caption{The misalignment direction is convergent across all layers (mean cosine
of per-seed directions with the seed-averaged direction), but cleanly separated
from benign training noise only in the early-middle layers, where the
benign-versus-benign null is small. The separation degrades with depth.}
\label{tab:mis-converge}
\end{table}

\paragraph{Where misalignment is linearly readable.}
The convergence is depth-stable near $0.97$, but the benign null climbs from
$0.16$ at layer 12 to $0.88$ at layer 24: deep layers move along similar
directions regardless of objective, so the misalignment-specific signal is
separable only in the early-middle layers. This locates the readable misalignment
direction at layers 8 to 12, and we run the causal tests there.

\paragraph{The direction is causally necessary for misalignment.}
Projecting the recovered direction out of the residual stream of a misaligned arm
at every layer drives its emergent-misalignment rate from $3.6\%$ to $0.0\%$,
while ablating a random direction of the same dimension leaves it unchanged at
$3.4\%$ (replicated across two runs). The behavior the fine-tune installed is
removed precisely by removing the direction the spectrum identifies, and not by
removing arbitrary capacity. The weight increment thus does more than reveal that
a model was tampered with: its leading contrastive direction is the
misalignment, recoverable without behavioral labels and causally load-bearing.
% SUFFICIENCY paragraph pending the fine-alpha (0.05-0.5) coherent-regime sweep
% (causal_misalign_fine.json); do not assert the null until that lands.
