\section{Related work}
\label{sec:related}

\paragraph{Spectra of trained networks.}
Heavy-tailed self-regularization reads model quality off the eigenvalue spectra
of trained weight matrices, finding that well-trained layers develop
heavy-tailed spectra with outliers above a Marchenko--Pastur bulk, and that a
``bulk plus spikes'' regime appears during training \citep{martin2021}. That
line analyzes the trained weight $W$ and predicts generalization; we analyze the
\emph{increment} $\dW$ between two checkpoints and ask what its spikes do
behaviorally. A recent random-matrix analysis shows that the \emph{small} singular directions
of the trained weight $W$ carry information that fine-tuning edits
\citep{staats2024small}; that statement is about $W$ itself, whereas we analyze
the increment $\dW$, whose \emph{leading} spikes are the edits, so the two
pictures describe different objects and are not in tension. We add the causal
link to a specific safety behavior.

\paragraph{Low-rank structure of fine-tuning.}
Fine-tuning has long been observed to occupy a low-dimensional subspace:
adaptation has low intrinsic dimension \citep{aghajanyan2021}, and low-rank
adapters recover most of full fine-tuning \citep{hu2022lora}. These are facts
about \emph{efficient training}. We show that the alignment increment of a
\emph{fully} fine-tuned released model is itself low rank in the spectral
sense, with the rank measured against a random-matrix null rather than assumed
by a parameterization. We do not use low rank alone as an alignment or
misalignment classifier. Its role is to make moved directions detectable and
locatable; the safety claim comes only after a matched contrast identifies which
direction moved and an intervention tests whether that direction is
ablation-sensitive.

\paragraph{Weight-space arithmetic.}
Task vectors edit behavior by adding or negating whole weight increments: the
difference of a fine-tuned and a base model is a direction in weight space, and
negating it removes a learned task while adding it installs one
\citep{ilharco2023task, ortizjimenez2023task}. Our misalignment direction is a
contrastive task vector, the difference of two arms' increments. Our ablation
and steering tests are activation-space analogues of negation and addition for
the recovered residual-stream direction, rather than whole-weight task-vector arithmetic. The
contribution is not the differencing but the spectral \emph{localization}: we
isolate the low-rank, spiked subspace of the increment against a random-matrix
threshold and show ablation sensitivity for the refusal subspace and for the
contrastive misalignment direction recovered across independent fine-tunes.

\paragraph{Directions for safety behavior.}
Refusal in chat models is mediated by a single residual-stream direction that
can be found by difference-of-means and removed by projection
\citep{arditi2024}, and representation-engineering methods read and steer
concept directions more generally \citep{zou2023repe}. Sparse autoencoders
surface interpretable, steerable features relevant to model behavior
\citep{templeton2026scaling}. These methods locate behavior in
\emph{activation} space using behavioral labels. We instead start from the
\emph{weight} increment, with no behavioral labels, and show that its top
singular subspace contains an enriched component of the refusal direction, tying
the label-free spectral object to the supervised one.

\paragraph{Supervised and audit baselines.}
Supervised linear probes can detect strategic deception when honest-versus-deceptive
labels are available \citep{goldowskydill2025}, and hidden-objective audits can
surface latent objectives when auditors have enough access to data, examples, and
investigation time \citep{marks2025}. These are stronger tools in the regime
where the relevant behavior is already represented in labeled examples or known
triggers. Our target regime is narrower: a same-base, same-recipe comparison in
which the direction is recovered from weight increments before using behavioral
examples to fit a probe.

\paragraph{Emergent misalignment.}
Narrow fine-tuning can induce broad misalignment \citep{betley2025emergent}, and
model-organism follow-up work makes the setting cleaner and more controllable
\citep{turner2025organisms}. \citet{soligo2025convergent} show the resulting
behavior is mediated by a convergent linear direction recovered from model
\emph{activations}, with the behavior present; \citet{turner2025organisms} show
that clean organisms can also be induced with rank-one adapters. We recover a convergent
misalignment direction as well, but from the differenced \emph{weight} increment
of matched arms, without using behavioral examples to fit it, and locate it via the increment's
spectrum; we also establish the same spectral picture for the intended
\emph{alignment} of a released model. Spectral signatures have also been used to detect backdoors from
poisoned-data representations \citep{tran2018spectral}; that detector operates
per input on activations, whereas ours operates per model on weights.
Layer-aware filtering work reports that safety-degrading fine-tuning can raise
the effective rank of inference-time activations on harmful prompts
\citep{larf2025}; this is a useful tension but a different object from the
weight-increment spectrum studied here, where concentrated causes can still
produce diffuse downstream activations.
