\documentclass[11pt]{article}

% =====================================================================
%  The Spectral Geometry of Misalignment
%  Formal theory document for the alignment-geometry project.
%  Standard packages only; compiles with tectonic or latexmk -pdf.
% =====================================================================

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{enumitem}
\usepackage[hidelinks]{hyperref}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\norm}[1]{\lVert #1 \rVert}
\newcommand{\inner}[2]{\langle #1, #2 \rangle}
\newcommand{\as}{\text{a.s.}}
\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\dW}{\Delta W}
\newcommand{\Wb}{W_{\mathrm{b}}}
\newcommand{\Wm}{W_{\mathrm{m}}}
\newcommand{\MP}{\mathrm{MP}}
\newcommand{\TW}{\mathrm{TW}}

\title{\bfseries The Spectral Geometry of Misalignment}
\author{Aryan Gupta\\ \texttt{aryan.cs.app@gmail.com}}
\date{June 2026}

\begin{document}
\maketitle

\begin{abstract}
\noindent
Fine-tuning a language model is a perturbation of its weights. We ask whether,
under a matched benign comparison, the \emph{spectrum} of that perturbation can
provide evidence for a concentrated update associated with a misaligned
objective, without using labeled examples of bad behavior to fit the statistic.
We model the weight increment $\dW = W_{\mathrm{ft}} - W_{\mathrm{base}}$ of a single
layer as an iid Gaussian rectangular matrix plus a deterministic fixed-rank signal,
and apply the additive rectangular singular-value transition of
Benaych-Georges and Nadakuditi \cite{benaych2012}. The classical
Baik--Ben~Arous--P\'ech\'e (BBP) spiked-covariance results
\cite{bbp2005,baik2006,paul2007} provide historical context but concern a
different, multiplicative Gaussian covariance model.
First, a planted signal direction of strength
$\theta$ produces an eigenvalue that detaches from the Marchenko--Pastur bulk
\cite{marchenko1967} if and only if $\theta > \sqrt{\gamma}$, where $\gamma$ is
the aspect ratio of the layer.
Second, and centrally, at \emph{fixed} signal energy $\tau$ and near-uniform energy
allocation, a perturbation is spectrally detectable exactly when its rank is below a
critical value $r_\star = \tau / \sqrt{\gamma}$: a concentrated update crosses the
threshold while an energy-matched diffuse update of higher rank stays invisible.
Rank, not magnitude alone, is the discriminator in that controlled comparison.
Third, above threshold the leading sample eigenvector aligns with
the planted direction with computable overlap, so the test also yields
a label-free estimate of the perturbation direction under the ideal model. We pair these
with a known-scale Tracy--Widom null \cite{tracywidom1996,johnstone2001}
calibrated for the Gaussian asymptotic model, so false-positive control applies
only under that ideal null; a fitted scale is an operational heuristic unless
recalibrated.
The theorems establish a
conditional: \emph{given} that a misaligned fine-tune deposits a low-rank signal and an
energy-matched benign fine-tune spreads its budget over higher rank, the spectral test
separates them. Whether real misaligned fine-tunes are in fact low-rank at matched
energy, while benign ones are diffuse, is an empirical claim, and measuring the
benign-side rank is the experiment on which everything rests. We give the construction,
the estimator, the confounds that can counterfeit a spike, and where the assumptions of
the theory fail for transformer activations.
\end{abstract}

\section{Introduction}
\label{sec:intro}

A practitioner is handed a fine-tuned language model and asked a hard question: does it
carry a hidden, misaligned objective? The strongest available white-box answers come
from supervised probes. Trained on labeled examples of a model lying or defecting, a
linear probe reads deception from the residual stream at very high accuracy
\cite{goldowskydill2025}. The difficulty is that the supervision is
exactly what one lacks. If one already possessed reliable labels for when a model
pursues a hidden goal, the detection problem would be largely solved. The binding
constraint in practice is access to such labels, and to the relevant distribution.
It would overstate the recent auditing evidence to say that the decisive resource in
surfacing a hidden objective was access to the training data: in one four-team game,
all three teams with model-and-data access succeeded while an API-only team did not,
but the access conditions were confounded and the authors note that explicit synthetic
documents made the task unusually favorable to data search \cite{marks2025}.

This motivates a different question. Fine-tuning is a transformation of the weights.
Rather than asking what the model says, we ask what the fine-tune \emph{did} to the
weight matrices, and whether that leaves a structural trace that a misaligned update
shares and a benign one does not. The object we examine is the weight increment
$\dW = W_{\mathrm{ft}} - W_{\mathrm{base}}$, and the tool is the spectrum of $\dW$. The
hope is a same-recipe screening statistic that can be computed before fitting a
behavioral probe, then calibrated against a matched benign control and validated
behaviorally.

This hope has precedent, and the precedents constrain what we can claim. Heavy-tailed
self-regularization reads a label-free, data-free fingerprint of model quality directly
off weight spectra, and its taxonomy already includes a ``bulk plus spikes'' regime
\cite{martin2021}. Spectral outlier analysis detects data poisoning from the singular
structure of learned representations \cite{tran2018}. Random-matrix tools have very
recently been turned on transformer internals for hallucination detection
\cite{ettori2026}, and a closely related line ties weight spectra,
activation covariance, and fine-tuning together on real language models
\cite{staats2024}. None of these targets misalignment, and none isolates the structural
question we pose. But they mean our contribution cannot be ``use spectra to inspect a
model''; that ground is taken. The contribution has to be sharper.

\paragraph{Contributions.}
We make the following.
\begin{enumerate}[leftmargin=1.5em,itemsep=2pt]
\item A random-matrix model of fine-tuning as a deterministic fixed-rank
perturbation of an iid Gaussian layer increment, with the diffuse part playing
the role of the Marchenko--Pastur bulk \cite{marchenko1967} and the additive
rectangular transition governed by \cite{benaych2012}
(Section~\ref{sec:model}).
\item A detectability theorem: a signal direction of strength $\theta$ is spectrally
visible exactly when $\theta > \sqrt{\gamma}$, with the eigenvalue location given in
closed form (Proposition~\ref{prop:detect}).
\item The central statement, the \emph{rank-at-fixed-energy} discriminator: at matched
signal energy and equal or diffuse energy allocation, an update is detectable
exactly when its rank is below $r_\star = \tau/\sqrt{\gamma}$, so two
energy-matched fine-tunes of concentrated and diffuse rank are spectrally separated
(Proposition~\ref{prop:rank}). We are not aware of prior work in the cited
spectral-fine-tuning literature that isolates this rank-at-fixed-energy
comparison; the prior results above instead conflate the magnitude of an update
with how that movement is structured.
\item A label-free identification result: above threshold the leading sample eigenvector
estimates the planted direction with computable overlap
(Proposition~\ref{prop:overlap}), and a known-scale Gaussian Tracy--Widom test
with ideal-null false-positive control (Proposition~\ref{prop:test}).
\item A subspace observable, the principal-angle distance between the leading eigenspaces
of matched models, intended to capture signal carried by eigenvector rotation rather than
eigenvalue magnitude (Definition~\ref{def:subspace}, Remark~\ref{rem:subspace}), with
surrogate calibration as a complement to scalar spectral summaries.
\item The limitations (Section~\ref{sec:limits}): which assumptions transformer
weights and activations violate, why the activation-covariance instrument is weaker than
the weight increment, a sign analysis reconciling an apparently contradictory empirical
result, and the experiment that the whole program stands or falls on.
\end{enumerate}

\paragraph{What is proved and what is assumed.}
The theorems
of Section~\ref{sec:results} prove an \emph{ideal-model implication}: if a misaligned fine-tune
deposits a rank-$r_{\mathrm m}$ signal and a benign fine-tune of equal energy spreads its
perturbation over a fixed rank $r_{\mathrm b} \gg r_{\mathrm m}$, then the spectral test
separates the two, with the threshold and the critical rank given explicitly. The
\emph{antecedent}, that misaligned updates are concentrated and benign updates of equal
energy are diffuse, is an empirical hypothesis. The supporting evidence is suggestive but
incomplete: emergent misalignment is mediated by a single, convergent linear direction
\cite{soligo2025}, a single rank-one adapter suffices to induce it \cite{turner2025}, and
emergent-misalignment weight deltas occupy a shared low-dimensional subspace across tasks
\cite{shared2025}. What we have not found in the cited prior work is a
matched-energy measurement of the benign side. That
measurement is the linchpin of the empirical program (Section~\ref{sec:program}), and the
theory is written so that it owns only the implication.

\paragraph{A note on the name.}
The project is named for an analogy with Fourier inspection of vision models, but the
mathematics here is spectral and random-matrix theoretic, not Fourier analytic. A weight
matrix carries no canonical periodic axis, so the discrete Fourier transform is not the
natural basis; the singular value decomposition is. We use the latter and reserve
genuine Fourier analysis for settings with group structure, where it is the correct
basis, which is outside the static-detection scope of this document.

\section{Setup and notation}
\label{sec:setup}

Fix a layer with weight matrix $W \in \R^{p \times q}$, and write $W_{\mathrm{base}}$ for
its value in a reference model and $W_{\mathrm{ft}}$ for its value after fine-tuning. The
object of study is the increment
\[
  \dW \;=\; W_{\mathrm{ft}} - W_{\mathrm{base}} \;\in\; \R^{p \times q}.
\]
Without loss of generality take $q \le p$ (otherwise transpose), and set the aspect ratio
\[
  \gamma \;=\; q/p \;\in\; (0,1].
\]
We treat the $p$ rows of $\dW$ as $p$ observations in $\R^q$ and form the symmetric
positive semidefinite matrix
\[
  C \;=\; \frac{1}{p}\, \dW^{\!\top} \dW \;\in\; \R^{q \times q}.
\]
The nonzero eigenvalues of $C$ are the squared singular values of $\dW$ divided by $p$,
so the spectrum of $C$ is a faithful and scale-controlled summary of the singular
structure of the update. We write $\lambda_1(C) \ge \cdots \ge \lambda_q(C) \ge 0$ for its
eigenvalues and $\hat v_1, \dots, \hat v_q$ for the corresponding unit eigenvectors.

\paragraph{Orientation.}
The covariance $C=\dW^{\!\top}\dW/p$ is input-side: its eigenvectors are the right
singular vectors of $\dW$. For residual-stream interventions in the empirical paper we
use the output-side covariance $\widetilde C=\dW\dW^{\!\top}/q$, whose eigenvectors are the
left singular vectors. Transposition preserves singular values and algebraic rank, but
the displayed covariance normalizations differ: the nonzero eigenvalues of
$\widetilde C$ are $p/q$ times those of $C$. Its null edge and fluctuation scale must
therefore be rescaled accordingly. The dimensionless additive detectability condition is
equivalent after transposition. We state all scalar formulas in the $q\le p$, input-side
normalization, while behavioral interventions use output-side singular vectors.

\paragraph{Energy.}
The \emph{energy} of the update is its squared Frobenius norm,
$\norm{\dW}_F^2 = \Tr(\dW^{\!\top}\dW) = p\,\Tr(C)$. Energy is a weight-space quantity, and
it is the natural budget to hold fixed when comparing two fine-tunes: a benign and a
misaligned update that move the weights by the same amount differ, if at all, in
\emph{how} they spend that budget across directions, not in how much they spend.

\paragraph{Matched organisms.}
The comparison the theory is built for involves three models sharing a base:
$W_{\mathrm{base}}$; a benign control $\Wb$ obtained by fine-tuning on an
alignment-neutral objective; and a misaligned model $\Wm$ obtained by an otherwise
identical recipe that installs the misaligned objective. Write
$\dW_{\mathrm b} = \Wb - W_{\mathrm{base}}$ and $\dW_{\mathrm m} = \Wm - W_{\mathrm{base}}$,
with associated $C_{\mathrm b}, C_{\mathrm m}$. The control is energy-matched,
$\norm{\dW_{\mathrm b}}_F = \norm{\dW_{\mathrm m}}_F$, so that any spectral difference is
attributable to the structure of the update and not to its magnitude. The need for this
control, rather than a comparison to the base model alone, is that \emph{any} fine-tune
moves the spectrum; a base-versus-misaligned difference would only certify that
fine-tuning occurred.

\section{Random-matrix preliminaries}
\label{sec:rmt}

We collect the results we use. They are standard; we state them in the normalization of
Section~\ref{sec:setup} and cite primary sources. Throughout, the asymptotic regime is
$p, q \to \infty$ with $q/p \to \gamma \in (0,1]$.

\subsection{The Gaussian bulk}

\begin{theorem}[Marchenko--Pastur law and Gaussian edge
\cite{marchenko1967,bai2010}]
\label{thm:mp}
Let $X \in \R^{p \times q}$ have iid $\mathcal N(0,\sigma^2)$ entries and let
$S = \tfrac1p X^{\!\top} X$. As $p,q \to \infty$ with $q/p \to \gamma
\in (0,1]$, the empirical spectral distribution of $S$ converges almost surely
to the Marchenko--Pastur law supported on $[\lambda_-,\lambda_+]$, where
\[
  \lambda_\pm=\sigma^2(1\pm\sqrt\gamma)^2.
\]
The largest eigenvalue also converges almost surely to
$\lambda_+=\sigma^2(1+\sqrt\gamma)^2$.
\end{theorem}

The upper edge is the asymptotic reference level for the ideal Gaussian null.
For finite matrices, an edge exceedance is descriptive until it is compared with
the null fluctuation scale of Theorem~\ref{thm:tw}.

\subsection{The additive rectangular spike: the primary model}

\begin{theorem}[Rectangular additive spike \cite{benaych2012}]
\label{thm:additive}
Let $N\in\R^{p\times q}$ have iid $\mathcal N(0,\sigma^2)$ entries, with
$q/p\to\gamma\in(0,1]$. Let $a\in\R^p$ and $b\in\R^q$ be deterministic unit
vectors, set $S=\beta ab^{\!\top}$, and define
$\theta=\beta^2/(p\sigma^2)$. For
$C=p^{-1}(N+S)^{\!\top}(N+S)$,
\[
  \lambda_1(C)\xrightarrow{\as}
  \begin{cases}
    \sigma^2(1+\theta)\left(1+\dfrac{\gamma}{\theta}\right),
      & \theta>\sqrt\gamma,\\[1.4ex]
    \sigma^2(1+\sqrt\gamma)^2, & \theta\le\sqrt\gamma.
  \end{cases}
\]
If $\hat b_1$ is the leading right singular vector of $N+S$, then
\[
  |\inner{\hat b_1}{b}|^2\xrightarrow{\as}
  \begin{cases}
    \dfrac{1-\gamma/\theta^2}{1+\gamma/\theta},
      & \theta>\sqrt\gamma,\\[1.4ex]
    0, & \theta\le\sqrt\gamma.
  \end{cases}
\]
\end{theorem}

This is the theorem used below. Its Gaussian specialization permits deterministic
signal directions because the noise is bi-orthogonally invariant. The cited general
theory also treats fixed-rank perturbations under explicit isotropy and edge
assumptions; it does not justify arbitrary growing-rank or correlated-noise updates.

\subsection{Historical covariance analogue}

The classical Baik--Ben~Arous--P\'ech\'e literature
\cite{bbp2005,baik2006,paul2007} studies a different model: a low-rank
perturbation of a population covariance \cite{johnstone2001}. The
following Gaussian, simple-spike statements explain the historical terminology,
but they are not used to turn the additive checkpoint delta into a sample
covariance.

\begin{theorem}[Classical spiked-covariance eigenvalue limit
\cite{baik2006}]
\label{thm:bbp}
Let the rows of $Y\in\R^{p\times q}$ be iid
$\mathcal N(0,\Sigma)$, where
$\Sigma=\sigma^2(I_q+\theta vv^{\!\top})$, $v$ is a deterministic unit vector,
and the population spike is simple. For $S_{\rm cov}=p^{-1}Y^{\!\top}Y$ and
$q/p\to\gamma\in(0,1]$,
\[
  \lambda_1(S_{\rm cov})\xrightarrow{\as}
  \begin{cases}
    \sigma^2(1+\theta)\left(1+\dfrac{\gamma}{\theta}\right),
      & \theta>\sqrt\gamma,\\[1.4ex]
    \sigma^2(1+\sqrt\gamma)^2, & \theta\le\sqrt\gamma.
  \end{cases}
\]
\end{theorem}

\begin{theorem}[Classical spiked-covariance eigenvector overlap
\cite{paul2007}]
\label{thm:overlap}
Under the Gaussian simple-spike conditions of Theorem~\ref{thm:bbp}, the
leading sample eigenvector $\hat v_1$ satisfies
\[
  |\inner{\hat v_1}{v}|^2\xrightarrow{\as}
  \begin{cases}
    \dfrac{1-\gamma/\theta^2}{1+\gamma/\theta},
      & \theta>\sqrt\gamma,\\[1.4ex]
    0, & \theta\le\sqrt\gamma.
  \end{cases}
\]
\end{theorem}

The coincidence of the white-noise formulas does not make the additive and
multiplicative models probabilistically equivalent.

\subsection{The known-scale null and a qualitative general bulk}

\begin{theorem}[Known-scale Gaussian Tracy--Widom edge
\cite{tracywidom1996,johnstone2001}]
\label{thm:tw}
For $S=p^{-1}N^{\!\top}N$ with iid real
$N_{ij}\sim\mathcal N(0,\sigma^2)$, known $\sigma^2$, and
$q/p\to\gamma\in(0,1]$,
\[
  \frac{\lambda_1(S)-\mu_{p,q}}{s_{p,q}}\Rightarrow\TW_1,
\]
where $\TW_1$ is the order-one Tracy--Widom law and
\[
  \mu_{p,q}=\frac{\sigma^2}{p}(\sqrt{p-1}+\sqrt q)^2,
  \qquad
  s_{p,q}=\frac{\sigma^2}{p}(\sqrt{p-1}+\sqrt q)
  \left(\frac1{\sqrt{p-1}}+\frac1{\sqrt q}\right)^{1/3}.
\]
The law was introduced for orthogonal ensembles by Tracy and Widom; Johnstone
established this real Wishart/PCA scaling.
\end{theorem}

\begin{remark}[Qualitative rectangular D-transform fallback]
\label{rem:rectangular-general}
For a non-white rectangular noise matrix with a compact limiting singular-value
law, no stray noise singular values, and the invariance or isotropic-direction
hypotheses of \cite{benaych2012}, a deterministic fixed-rank perturbation has a
qualitative outlier transition governed by the rectangular $D$-transform of the
noise singular-value law. The outlier location and threshold must be computed
from that law; the Marchenko--Pastur closed forms above do not carry over. Real
checkpoint increments need not satisfy the theorem's invariance assumptions, so
empirical bulk and surrogate checks remain necessary.
\end{remark}

\section{A spiked model of fine-tuning}
\label{sec:model}

We now place the weight increment inside the spiked framework. The modeling content is
a decomposition of the update into a diffuse part and a structured part, together with
the assumptions under which the diffuse part behaves as a Marchenko--Pastur bulk
\cite{marchenko1967}.

\begin{definition}[Diffuse-plus-signal decomposition]
\label{def:decomp}
Write the increment as
\[
  \dW \;=\; N \;+\; S, \qquad
  S \;=\; \sum_{i=1}^{r} \beta_i\, a_i b_i^{\!\top},
\]
where $N \in \R^{p\times q}$ is the diffuse component and $S$ is deterministic
with fixed rank $r=\rank(S)$ as $p,q\to\infty$. The vectors
$a_i\in\R^p$ and $b_i\in\R^q$ are orthonormal families and
$\beta_1\ge\cdots\ge\beta_r>0$. The dimensionless additive spike strengths are
$\theta_i=\beta_i^2/(p\sigma^2)$, and we consider sequences on which these
strengths have fixed positive limits. Here $\sigma^2$ is the per-entry variance
of $N$ from Assumption~\ref{ass:bulk}.
\end{definition}

\begin{assumption}[Ideal Gaussian diffuse bulk]
\label{ass:bulk}
The diffuse component has iid entries $N_{jk}\sim\mathcal N(0,\sigma^2)$.
\end{assumption}

Assumption~\ref{ass:bulk} is the load-bearing modeling choice, and the reason we work
with the increment $\dW$ rather than the raw weight $W$. Many trained neural-network
weight spectra are empirically heavy-tailed or otherwise depart from a pure
Marchenko--Pastur law \cite{martin2021}, so a white-noise null on $W$ is often
mis-specified. This does not establish that a fine-tune increment is Gaussian or
isotropic. The increment makes that hypothesis directly testable, and the closed forms
of Theorems~\ref{thm:mp}, \ref{thm:additive}, and \ref{thm:tw} are used only when the
ideal assumptions are imposed. Remark~\ref{rem:rectangular-general} describes the more
cautious qualitative alternative when the bulk is non-white.

\begin{assumption}[Structure hypothesis, to be tested]
\label{ass:structure}
A misalignment-relevant fine-tune deposits its signal in low rank, $r_{\mathrm m}$ small,
while a benign fine-tune of equal energy spreads a comparable Frobenius budget over higher
rank, $r_{\mathrm b} \gg r_{\mathrm m}$.
\end{assumption}

Assumption~\ref{ass:structure} is not proved here and is not folklore. It is the empirical
hypothesis the program tests, and the theorems below deliver their conclusion
\emph{conditional} on it.

\begin{remark}[Why energy is the right thing to fix]
\label{rem:energy}
By Definition~\ref{def:decomp}, the signal contributes
$\norm{S}_F^2 = \sum_i \beta_i^2 = p\,\sigma^2 \sum_i \theta_i$ to the energy. Holding the
signal energy fixed is therefore holding $\tau := \sum_i \theta_i$ fixed. The discriminator
of Section~\ref{sec:results} is precisely a statement about how a fixed $\tau$ is
distributed across $r$ directions, which is why energy matching is built into the matched
organisms of Section~\ref{sec:setup}.
\end{remark}

\section{Main results}
\label{sec:results}

\subsection{Detectability}

\begin{proposition}[Detectability and location of a single spike]
\label{prop:detect}
Under Assumption~\ref{ass:bulk} and the decomposition of Definition~\ref{def:decomp} with
a single signal direction of induced strength $\theta$, the leading eigenvalue of
$C = \tfrac1p \dW^{\!\top}\dW$ satisfies, as $p,q\to\infty$ with $q/p\to\gamma$,
\[
  \lambda_1(C) \;\xrightarrow{\as}\;
  \begin{cases}
    \sigma^2(1+\theta)\!\left(1+\dfrac{\gamma}{\theta}\right), & \theta > \sqrt{\gamma},\\[1.4ex]
    \sigma^2(1+\sqrt\gamma)^2, & \theta \le \sqrt{\gamma}.
  \end{cases}
\]
Consequently the signal is asymptotically visible in the spectrum, that is
$\lambda_1(C)$ exceeds the bulk edge by a deterministic gap, if and only if $\theta >
\sqrt{\gamma}$.
\end{proposition}

\begin{proof}
This is Theorem~\ref{thm:additive} with
$\theta=\beta^2/(p\sigma^2)$. No population-covariance equivalence is used.
\end{proof}

\subsection{The discriminator: rank at fixed energy}

For deterministic fixed $r$, the rectangular finite-rank theorem applies to each
planted singular direction. The detection of \emph{any} outlier is governed by
$\max_i\theta_i$, because the leading singular value tracks the strongest direction.

\begin{proposition}[Rank-at-fixed-energy discriminator]
\label{prop:rank}
For a fixed rank $r$ in the proportional limit, fix the signal energy through
$\tau = \sum_{i=1}^r \theta_i$. Among rank-$r$ signals of
this energy, the leading induced strength satisfies $\max_i \theta_i \ge \tau/r$, with
equality when the energy is spread equally, $\theta_i = \tau/r$. Therefore:
\begin{enumerate}[leftmargin=1.6em,itemsep=2pt]
\item An equal-energy, equal-strength rank-$r$ update is spectrally detectable
(Proposition~\ref{prop:detect}) if and only if
\[
  \frac{\tau}{r} > \sqrt{\gamma}
  \quad\Longleftrightarrow\quad
  r < r_\star, \qquad r_\star := \frac{\tau}{\sqrt{\gamma}}.
\]
\item Consequently, a concentrated update with $r_{\mathrm m} < r_\star$ produces a
supercritical spike above the bulk edge, while an energy-matched \emph{diffuse} update that
spreads its budget approximately uniformly over rank $r_{\mathrm b} \ge r_\star$ keeps every
direction subcritical and is asymptotically invisible. The test separates a concentrated
update from a diffuse one of equal energy; a high-rank update that nonetheless places a
supercritical share of its energy in a single direction is still detected.
\end{enumerate}
\end{proposition}

\begin{proof}
The inequality $\max_i \theta_i \ge \tau/r$ is the pigeonhole bound on a sum of $r$
nonnegative terms, with equality at the uniform allocation $\theta_i = \tau/r$. For the
uniform allocation, every spike has strength $\tau/r$, so by Proposition~\ref{prop:detect}
a detached eigenvalue exists if and only if $\tau/r > \sqrt\gamma$, equivalently $r <
\tau/\sqrt\gamma = r_\star$. For statement (2): when $r_{\mathrm m} < r_\star$ the leading
strength exceeds $\sqrt\gamma$ and a spike appears, while a uniform allocation over
$r_{\mathrm b} \ge r_\star$ puts every strength at or below $\sqrt\gamma$ and no eigenvalue
detaches. The pigeonhole bound is one-sided, so invisibility requires the diffuse,
near-uniform allocation rather than $r_{\mathrm b} \ge r_\star$ alone; a high-rank update
concentrated in one direction has $\max_i \theta_i > \sqrt\gamma$ and is detected.
\end{proof}

Proposition~\ref{prop:rank} is the conceptual core. At fixed magnitude, detectability is
governed by rank: a concentrated update is supercritical, an energy-matched diffuse update
is not, and the boundary is the critical rank $r_\star = \tau/\sqrt\gamma$. Conditional on
Assumption~\ref{ass:structure}, which assigns the misaligned fine-tune the low rank and the
benign one the high rank, this is a label-free separation. The proposition proves the
implication, and the experiment of
Section~\ref{sec:program} tests whether the ranks fall on the asserted sides of
$r_\star$.

\begin{remark}[The uniform allocation is the benign best case]
\label{rem:uniform}
Spreading energy uniformly is the allocation that minimizes the leading strength, hence
the hardest case to detect. A benign update need not be uniform, but if even the uniform
allocation of its energy would be detectable then so is the update; the proposition states
the favorable-to-benign boundary. The substantive empirical question is whether benign
fine-tuning is diffuse \emph{enough}, that is whether $r_{\mathrm b} \ge r_\star$, not
whether it is exactly uniform.
\end{remark}

\subsection{Critical rank for representative layers}
\label{sec:layers}

The critical rank $r_\star = \tau/\sqrt\gamma$ depends on the layer through its aspect
ratio $\gamma$ and on the fine-tune through the signal energy $\tau$. Table~\ref{tab:rstar}
reports $\gamma$ and
$r_\star$ for the released Llama-3-8B architecture (hidden width $4096$,
feed-forward width $14336$, 32 query heads, and eight key/value heads)
\cite{grattafiori2024}, at two illustrative signal energies.

\begin{table}[h]
\centering
\begin{tabular}{lccc}
\hline
Layer & shape $(p \times q)$ & $\gamma = q/p$ & $r_\star$ at $\tau = 4,\ 16$ \\
\hline
Attention $q/o$ projection & $4096 \times 4096$ & $1.00$ & $4.0,\ \ 16.0$ \\
Attention $k/v$ projection & $4096 \times 1024$ & $0.25$ & $8.0,\ \ 32.0$ \\
MLP up / gate projection & $14336 \times 4096$ & $0.286$ & $7.5,\ \ 29.9$ \\
MLP down projection & $14336 \times 4096$ & $0.286$ & $7.5,\ \ 29.9$ \\
\hline
\end{tabular}
\caption{Critical rank $r_\star = \tau/\sqrt\gamma$ across layer shapes. Shapes are written
as (rows $\times$ columns) after the transpose of Section~\ref{sec:setup} that places
$q \le p$, so $\gamma = q/p$ is the smaller dimension over the larger. An
equal-strength rank-$r$ signal is detectable in a layer exactly when
$r < r_\star$ for that layer.}
\label{tab:rstar}
\end{table}

First, the rectangular key/value and feed-forward maps, with smaller $\gamma$,
tolerate a higher fixed-rank signal before it disappears into the bulk, while
the square query/output maps impose the strictest concentration requirement.
Second, the rank-one adapter that suffices
to induce emergent misalignment \cite{turner2025} sits at $r = 1$, far below $r_\star$ in
every layer, so the idealized equal-energy signal is well above threshold, whereas a
diffuse benign update would need to spread the same energy over at least $r_\star$
directions to stay hidden. The estimator of
Section~\ref{sec:estimator} therefore scans layers and reports the per-layer standardized
leading eigenvalue, with the prior that a signal, if present, is clearest where $\gamma$ is
small and the layer's signal energy is large.

\subsection{Label-free identification}

Under the ideal model and its test calibration, detection establishes that a spike
exists; under the same model, the eigenvector
estimates its planted direction.

\begin{proposition}[Label-free direction estimate]
\label{prop:overlap}
Under the conditions of Proposition~\ref{prop:detect} with $\theta > \sqrt\gamma$, the
leading eigenvector $\hat v_1$ of $C$ satisfies
\[
  |\inner{\hat v_1}{b}|^2 \;\xrightarrow{\as}\; \frac{1-\gamma/\theta^2}{1+\gamma/\theta} \;>\; 0,
\]
where $b$ is the planted signal direction. Thus $\hat v_1$ is asymptotically
nontrivially correlated with the perturbation direction, but is not a consistent
direction estimator at fixed $\theta$ and $\gamma$.
\end{proposition}

\begin{proof}
Immediate from the right-singular-vector statement of
Theorem~\ref{thm:additive}.
\end{proof}

Proposition~\ref{prop:overlap} matters for validation. If the recovered direction
$\hat v_1$ tracks the misalignment-relevant change, then steering the model along
$\pm \hat v_1$ should modulate the misaligned behavior, a causal check that the
spectral signal is behaviorally coupled rather than only a spectral artifact. This connects the spectral test to the established finding
that emergent misalignment is mediated by a convergent activation direction
\cite{soligo2025}, with the difference that here the direction is obtained from the
spectrum without alignment labels. Persona-vector work is related as a broader
example of low-dimensional control of model traits \cite{chen2025}, but is not an
emergent-misalignment direction result.

\subsection{A calibrated test}

\begin{proposition}[Known-scale level-$\alpha$ edge test]
\label{prop:test}
Let $H_0$ be the pure-noise null, $\dW = N$ with no signal (Assumption~\ref{ass:bulk} with
$\theta = 0$), and suppose $\sigma^2$ is known. Using the centering and scaling of
Theorem~\ref{thm:tw}, the test that rejects $H_0$ when
\[
  \frac{\lambda_1(C) - \mu_{p,q}}{s_{p,q}} \;>\; t_{1-\alpha},
  \qquad t_{1-\alpha} = \TW_1\text{ quantile},
\]
has asymptotic false-positive rate $\alpha$. Null eigenvalue fluctuations have
scale $p^{-2/3}$, while the critical window in spike strength around
$\theta=\sqrt\gamma$ has scale $p^{-1/3}$.
\end{proposition}

\begin{proof}
Under $H_0$ the leading eigenvalue obeys the Tracy--Widom limit of Theorem~\ref{thm:tw},
so the standardized statistic exceeds $t_{1-\alpha}$ with probability $\alpha$ in the
limit. For $\theta=\sqrt\gamma+\epsilon$, expansion of the outlier location in
Proposition~\ref{prop:detect} gives
\[
  \sigma^2(1+\theta)(1+\gamma/\theta)-\lambda_+
  =\frac{\sigma^2}{\sqrt\gamma}\epsilon^2+O(\epsilon^3).
\]
Matching this deterministic gap to the $p^{-2/3}$ edge scale gives
$\epsilon\asymp p^{-1/3}$. Inside that critical window, the pure-null
Tracy--Widom approximation does not by itself furnish a calibrated power
calculation.
\end{proof}

Proposition~\ref{prop:test} supplies the missing piece that purely descriptive spectral
statistics lack under the ideal Gaussian/null assumptions: a decision rule with a
controlled error rate and a stated resolution, derived from the null rather than
from a labeled validation set. Empirical screening still needs matched benign
calibration and behavioral validation.

\subsection{A subspace observable for rotation}

Scalar spectral statistics, such as effective rank or the leading eigenvalue, are
functions of the eigenvalues alone. It has been suggested that post-training reshapes
representation geometry in ways a magnitude test need not capture \cite{li2025geometry}, so we
add a complementary observable on the eigenvectors.

\begin{definition}[Leading-subspace principal-angle distance
\cite{edelman1998}]
\label{def:subspace}
For two models with covariances $C, C'$, let $V_k, V_k'$ be the matrices of their top-$k$
eigenvectors and let $\rho_1 \ge \cdots \ge \rho_k$ be the singular values of $V_k^{\!\top}
V_k'$, equal to the cosines of the principal angles between the two subspaces. Define the
Grassmann distance
\[
  d_k(C, C') \;=\; \Big( \sum_{j=1}^k \arccos^2(\rho_j) \Big)^{1/2}.
\]
\end{definition}

\begin{remark}[Calibrating the rotation observable]
\label{rem:subspace}
The distance $d_k$ can reveal rotation that an eigenvalue summary misses, but it
does not inherit Proposition~\ref{prop:test}'s scalar null. Under two independent
isotropic Gaussian null draws, the eigenvector bases are Haar-distributed
\cite{bai2010} and
independent leading directions are nearly orthogonal in high dimension. Matched
fine-tunes are correlated, however, so that observation does not define their null.
We therefore compare the candidate with its energy-matched benign control and use
design-specific resampling only as the surrogate sensitivity analysis of
Section~\ref{sec:permutation}. Any reported tail fraction is descriptive unless the
resampling transformation is justified by an exchangeable null. An asymptotic theory
of $d_k$ under a joint model of matched increments is left to future work.
\end{remark}

The subspace observable is label-free in the same sense as the others: it compares two
models, not a model against labeled behavior. It is the natural instrument when the
discriminator of interest is reorientation of existing capacity rather than addition of
new energy.

\section{From theory to a label-free estimator}
\label{sec:estimator}

The propositions describe population limits. We now specify the finite-sample procedure,
the secondary instrument on activations, and the confounds that can counterfeit a spike.

\subsection{The procedure}

\begin{enumerate}[leftmargin=1.6em,itemsep=3pt]
\item \textbf{Form the increment.} For each layer of interest compute $\dW = W_{\mathrm{ft}}
- W_{\mathrm{base}}$ and the covariance $C = \tfrac1p \dW^{\!\top}\dW$. Record the aspect
ratio $\gamma = q/p$.
\item \textbf{Estimate the bulk.} Fit the noise level $\sigma^2$ from the bulk of the
spectrum (for example by matching the median eigenvalue to the Marchenko--Pastur median),
not from the full trace, so that a genuine outlier does not inflate the estimated floor.
This fitted scale is an operational estimate, not the known scale assumed by
Proposition~\ref{prop:test}.
\item \textbf{Screen for an edge excursion.} Report the plug-in standardized leading
eigenvalue as a heuristic score, the ideal-model implied strength $\hat\theta$ from
Proposition~\ref{prop:detect}, and the descriptive number of MP-edge exceedances. Do not
interpret the exceedance count as an estimate of algebraic signal rank. A calibrated
decision requires a known scale or a separate validation of the plug-in null.
\item \textbf{Match energy.} When comparing a misaligned candidate to a benign control,
rescale so $\norm{\dW_{\mathrm m}}_F = \norm{\dW_{\mathrm b}}_F$ before comparing spectra,
so the comparison is at fixed budget as Proposition~\ref{prop:rank} requires.
\item \textbf{Recover and test the direction.} Take $\hat v_1$ from
Proposition~\ref{prop:overlap}, map it back to a residual-stream direction, and test
causal coupling by steering or ablation. A direction that modulates the measured
behavior is evidence that the spike is behaviorally relevant, not that it exhausts
the mechanism.
\item \textbf{Check rotation.} Compute the subspace distance of
Definition~\ref{def:subspace} between matched models, to catch signal carried by
reorientation rather than magnitude.
\end{enumerate}

\subsection{Finite-size surrogate stress tests}
\label{sec:permutation}

Proposition~\ref{prop:test} is an asymptotic, known-scale Gaussian result. Real
increments are finite, non-Gaussian, correlated, and analyzed with a fitted scale, so
the plug-in Tracy--Widom statistic is a visibility heuristic rather than a calibrated
test. As a sensitivity check, one may independently permute all entries of $\dW$;
this preserves the global empirical entry multiset while destroying row and column
structure. Independent sign flips preserve magnitudes but preserve the signed marginal
only under a justified symmetric-sign null. For $B$ surrogates, report the corrected
tail fraction
\[
  \frac{1+\#\{b:\lambda_1(C^{(b)})\ge\lambda_1(C^{\rm obs})\}}{B+1}.
\]
Call this quantity a randomization $p$-value only when the chosen transformations are
exchangeable under a stated null; otherwise call it a surrogate tail fraction. Neither
transformation preserves optimizer, architectural, row, or column correlations. The
committed Llama-3-8B census is therefore described as MP-visible structure, not as an
empirical-null-confirmed outlier census.

\subsection{The activation-covariance instrument, and why it is secondary}
\label{sec:activation}

One can run the same machinery on the activation covariance
$\Sigma_{\mathrm{act}} = \tfrac1n \sum_{t} h_t h_t^{\!\top}$ of residual-stream states
$h_t$ over $n$ probe tokens, comparing matched models. This is attractive because the
causally relevant misalignment direction has been identified in activation space; in
one reported rank-one-LoRA experiment, the layer-24 adapter $B$ vector and a
mean-difference activation direction have cosine near $0.04$, yet ablating that
activation direction substantially suppresses the measured behavior \cite{soligo2025}. The activation
covariance is therefore the right object for \emph{identification}.

For \emph{detection}, however, it is the weaker substrate, and we treat it as a heuristic
secondary instrument rather than a setting where Propositions~\ref{prop:detect}--\ref{prop:test}
hold as stated. The reason is that the activation covariance is strongly anisotropic even
before any fine-tuning: a handful of directions dominate, and a small number of coordinates
carry outsized variance \cite{sun2024,gao2019,ethayarajh2019}. The isotropic-bulk
Assumption~\ref{ass:bulk} fails, so the closed forms of
Theorem~\ref{thm:additive} no longer apply. A generalized spiked population model
for a non-identity base covariance \cite{baiyao2012} is one possible asymptotic
analogue, but its independent-sample assumptions do not automatically hold for
correlated token states. The clean criterion ``crosses the Marchenko--Pastur edge'' becomes
``separates from an estimated deformed edge,'' which is both weaker and more fragile. The
weight-increment analysis makes the isotropy hypothesis directly testable; it does
not make that hypothesis true for activation covariance or for real updates.

\subsection{A sign analysis: reconciling an apparently opposite result}
\label{sec:larf}

A recent result compares models fine-tuned on the top- versus bottom-ranked subsets
of nominally benign datasets and reports higher effective rank for the top-ranked,
safety-degrading subset when processing harmful prompts \cite{larf2025}. Read naively,
``misaligned means higher rank'' is the
opposite of our ``misaligned means a low-rank spike.''
The tension dissolves once the object is fixed.

Effective rank \cite{roy2007} is $\exp(H)$ where $H = -\sum_i \pi_i \log \pi_i$ is the entropy of the
normalized spectrum $\pi_i = \lambda_i / \sum_j \lambda_j$. Adding a single dominant spike
to a covariance concentrates the distribution $\pi$, lowers $H$, and therefore
\emph{lowers} effective rank. So our model predicts that a low-rank misalignment signal,
seen in the relevant covariance, lowers effective rank, which is what one would expect of a
concentrated perturbation.

The reported result concerns a different matrix. It measures the effective rank of the
inference-time activation matrix on harmful prompts, in a model that was fine-tuned on
harmful \emph{content}, with no energy control. Our prediction concerns the rank of the
weight increment $\dW$, the cause, not the diversity of downstream activations on
triggering inputs, the effect. A low-rank cause is entirely compatible with a higher-rank
effect: a concentrated update can broaden the range of behaviors the model expresses on
the inputs that excite it. The two measurements are not the same quantity and do not have
to share a sign. We therefore commit to reporting both, the descriptive
weight-increment MP-edge exceedance count and
the inference-time activation effective rank, and we do not claim more reconciliation than
this sign analysis supports.

\subsection{Confounds that counterfeit a spike}
\label{sec:confounds}

Several artifacts can produce a leading-eigenvalue excursion that is not a misalignment
signal. The estimator must control each.
\begin{itemize}[leftmargin=1.4em,itemsep=2pt]
\item \textbf{Outlier coordinates.} A few residual-stream coordinates with very large
variance \cite{sun2024} create a top eigenvalue unrelated to any fine-tune. Control by
per-coordinate standardization or robust covariance before spectral analysis, and by
working on $\dW$ where the effect is weaker.
\item \textbf{Energy leakage.} An unmatched comparison detects ``trained more,'' not
``trained misaligned.'' Control by the energy matching of step~4.
\item \textbf{Aspect-ratio regime.} For very wide matrices or many probe tokens, $\gamma
\to 0$, the bulk collapses to a point and \emph{every} spike is nominally supercritical
($\sqrt\gamma \to 0$). In that regime detectability is trivial and the benign-versus-
misaligned separation rests entirely on the rank argument of
Proposition~\ref{prop:rank}, not on the threshold. Report $\gamma$ explicitly and interpret
the threshold only where it is nontrivial.
\item \textbf{Heavy-tailed bulk.} Many trained raw-weight spectra depart from a
pure Marchenko--Pastur law \cite{martin2021}. The Gaussian model is therefore
posed on $\dW$, but its bulk fit must still be checked. If it deviates, discard
the MP closed form and use Remark~\ref{rem:rectangular-general} only as a
qualitative guide.
\end{itemize}

\section{The empirical program in brief}
\label{sec:program}

The theory makes the experiment unambiguous. We state only the commitments the theory
forces; the full protocol is the project plan.

\paragraph{The linchpin.} Measure the rank of $\dW$ at matched energy for a benign control
and a misaligned model sharing a base, and locate both relative to $r_\star =
\tau/\sqrt\gamma$. Assumption~\ref{ass:structure} predicts $r_{\mathrm m} < r_\star \le
r_{\mathrm b}$. If instead the benign control is also low-rank at matched energy, the
separation of Proposition~\ref{prop:rank} is empty and the approach fails for that
organism. This single measurement is the one the program cannot avoid, and we are
not aware of prior work above that reports it for a matched
benign-versus-misaligned organism.

\paragraph{Substrate.} The matched comparison requires \emph{full} fine-tuning, not
low-rank adaptation. A rank-constrained adapter fixes the rank of $\dW$ by construction, so
it cannot exhibit a rank difference, and its increment has no bulk for a spike to cross.
Emergent-misalignment organisms exist as full fine-tunes and as rank-one adapters
\cite{betley2025,turner2025}; the rank-one adapters are useful for the identification and
steering checks of Proposition~\ref{prop:overlap}, but the detection claim of
Propositions~\ref{prop:detect}--\ref{prop:rank} must be evaluated on full fine-tunes with a
matched benign control trained under an identical recipe.

\paragraph{Prediction.} Under the rank-structure antecedent of
Assumption~\ref{ass:structure}, the ideal scalar prediction is that the misaligned model
shows a supercritical leading eigenvalue in $\dW$ at a layer where the benign control does
not. In real organisms this scalar test may fail because benign fine-tuning is also
anisotropic; then the residual evidence must come from the contrastive singular-vector
comparison, behavioral ablation, and calibrated same-recipe held-out screens rather than
from an edge-exceedance count alone. Where the antecedent fails, one or more of these measurements do
not separate, and the theory says which assumption was responsible.

\section{Assumptions, limitations, and threats to validity}
\label{sec:limits}

We list the ways the program can be wrong, beyond the confounds of
Section~\ref{sec:confounds}.

\begin{enumerate}[leftmargin=1.6em,itemsep=3pt]
\item \textbf{The antecedent may be false.} Assumption~\ref{ass:structure} is the bet. If
misaligned and benign fine-tunes are equally concentrated at matched energy, rank does not
discriminate and the method has no signal. This is measurable, and measuring it is the
linchpin.
\item \textbf{Anisotropy on activations.} The Gaussian isotropic-bulk assumption
is an explicit idealization for $\dW$ and is visibly inappropriate for activation
covariance, which is why the latter is secondary (Section~\ref{sec:activation}).
A deformed-bulk model is only a candidate there and is not developed in this document.
\item \textbf{Heavy tails on raw weights.} We choose to pose the
Marchenko--Pastur null on the diffuse part of the increment. Evidence that many
trained raw-weight spectra depart from MP \cite{martin2021} motivates not applying
the threshold directly to $W$; it does not prove that increments satisfy the null.
\item \textbf{Narrow versus broad misalignment.} Narrow fine-tuning can induce
broad behavioral misalignment \cite{betley2025}. Whether such broad behavior is
associated with a more concentrated weight update than narrowly scoped behavior
is not established by that source and remains an empirical question.
\item \textbf{Deception is not misalignment.} Probes that read a gap between a model's
internal belief and its statement catch instrumental liars but largely miss models with an
internalized objective, which present no such gap \cite{fanatics2026}. Our target is the
internalized case, which is the regime where behavioral probes are weakest; this is the
method's positioning, and it also means the method cannot borrow a belief-gap signal. The
spectral signal, if it exists, is structural, present in the weights whether or not the
model ever ``intends'' to deceive.
\item \textbf{Evasion.} A spectral monitor is still a monitor. A model trained against it,
or with knowledge of it, may be able to spread its misalignment update over higher rank to
stay below threshold, the precise move the rank discriminator describes, or to push signal
into a benign-looking subspace, as latent-space monitors have been evaded elsewhere
\cite{bailey2024}. We claim no robustness to an adaptive adversary and treat robustness as
a hypothesis to be tested against such an adversary, not a property of the method.
\item \textbf{Finite-size and regime gaps.} The limits are asymptotic. Real layers have
finite $p,q$, the increment is one ``sample'' in a sense the covariance construction has to
make precise, and the Tracy--Widom calibration assumes Gaussianity and a known
scale. These conditions fail for the fitted-scale empirical screen, which must be
checked by finite-matrix simulation and reported as heuristic unless separately calibrated.
\end{enumerate}

\section{Related work}
\label{sec:related}

\paragraph{Spectral fingerprints of models.} Heavy-tailed self-regularization reads model
quality from weight spectra without data or labels, and names a ``bulk plus spikes'' phase
\cite{martin2021}. Our framing inherits its vocabulary and turns it to a different target,
misalignment, with a different object, the increment, and a different theory, the
rectangular additive threshold rather than a power-law tail index. The closest prior work ties weight spectra,
activation covariance, and fine-tuning together on transformer language models and finds
that fine-tuning refines the model in the small-singular-value directions
\cite{staats2024}; we differ in asking for a benign-versus-misaligned contrast at matched
energy, which that work does not consider. Spectral fingerprints have also been used for
model lineage in a way deliberately invariant to fine-tuning \cite{ghost2026}, the
opposite use to ours. Random-matrix descriptions of network spectra have precedent in the
analysis of loss-surface Hessians \cite{pennington2017} and in the local spectral
statistics of trained weight matrices, which stay random-matrix-like even where the global
density is not \cite{thamm2022}.

\paragraph{Spectral and random-matrix detection.} Spectral outlier scores detect data
poisoning from learned representations, per input rather than per model \cite{tran2018}.
Random-matrix statistics have been applied to hallucination detection in language
models, again per input \cite{ettori2026}. Neither addresses a
per-model misalignment screening statistic, and neither uses the
rank-at-fixed-energy argument.

\paragraph{The structure of emergent misalignment.} Emergent misalignment, in which narrow
fine-tuning induces broad misalignment \cite{betley2025}, is mediated by a single
convergent linear direction in activations \cite{soligo2025}, can be induced in clean
model organisms by a rank-one adapter \cite{turner2025}, is affected by persona
features \cite{wang2025persona,chen2025}, and has reported shared low-dimensional
parameter subspaces across tasks \cite{shared2025}. This body of work motivates,
but does not establish, the weight-increment rank structure assumed here; our
contribution is to state that assumption explicitly, model its consequence as a
rectangular additive outlier transition, and make rank at fixed energy the discriminator.
The single-direction phenomenon is part of a broader pattern in which
safety-relevant behaviors can be low-dimensional, as for refusal
\cite{arditi2024}; fine-tuning objectives can be optimized in low-intrinsic-dimensional
parameter subspaces \cite{aghajanyan2021}, and explicitly low-rank adapters are effective
\cite{hu2022}.

\paragraph{Geometry of fine-tuning and safety.} A complementary line analyzes how
fine-tuning breaks safety through low-rank curvature subspaces and argues that benign tasks
can also degrade alignment \cite{springer2026}; this is a foil for the clean separation we
hypothesize and a reason to take the benign-side measurement seriously. The
representation-geometry of post-training has been mapped with effective-rank and
spectral-slope observables, establishing that post-training changes geometry
\cite{li2025geometry}; we position our claim narrowly as detection in matched models, not
the more general and now-established statement that post-training moves the spectrum.

\paragraph{Behavioural and label-based detection.} Supervised probes detect deception at
high accuracy when labels and the distribution are available
\cite{goldowskydill2025}, and interpretability-driven audits surface hidden
objectives given sufficient access \cite{marks2025}. Unsupervised methods that search for
latent knowledge exist \cite{burns2023}, but arbitrary features can satisfy their
consistency constraints \cite{farquhar2023}. Our method is complementary: it targets the label-free,
unknown-distribution regime where the supervised results do not apply, and where labels
exist and the distribution is known a probe should be expected to do better.

\section{Conclusion}
\label{sec:conclusion}

We have given a random-matrix account of when fine-tuning is spectrally visible, and used
it to isolate a single sharp claim: at controlled energy, rank discriminates a concentrated
update from a diffuse one, with an explicit critical rank $r_\star = \tau/\sqrt\gamma$ and a
known-scale Gaussian ideal-model test. The theorems prove that \emph{if}, within
the fixed-rank additive model, misaligned fine-tunes are low-rank and
energy-matched benign fine-tunes are diffuse, the spectral statistic separates them and estimates
a candidate planted perturbation direction without behavioral labels. The hypothesis that real fine-tunes fall on the
asserted sides of $r_\star$ is the one the empirical program must establish, and the
benign-side rank at matched energy is the measurement on which the entire approach rests.
We have converted a vague hope, that spectra reveal misalignment,
into a falsifiable conditional with a named experiment that can refute it.

\small
\begin{thebibliography}{99}

\bibitem{marchenko1967}
V.~A.~Mar\v{c}enko and L.~A.~Pastur,
``Distribution of eigenvalues for some sets of random matrices,''
\emph{Math.\ USSR-Sbornik} \textbf{1}(4):457--483, 1967.

\bibitem{bai2010}
Z.~Bai and J.~W.~Silverstein,
\emph{Spectral Analysis of Large Dimensional Random Matrices}, 2nd ed.,
Springer, 2010.

\bibitem{edelman1998}
A.~Edelman, T.~A.~Arias, and S.~T.~Smith,
``The geometry of algorithms with orthogonality constraints,''
\emph{SIAM J. Matrix Anal. Appl.} \textbf{20}(2):303--353, 1998.
\href{https://doi.org/10.1137/S0895479895290954}{doi:10.1137/S0895479895290954}.

\bibitem{johnstone2001}
I.~M.~Johnstone,
``On the distribution of the largest eigenvalue in principal components analysis,''
\emph{Ann.\ Statist.} \textbf{29}(2):295--327, 2001.

\bibitem{bbp2005}
J.~Baik, G.~Ben~Arous, and S.~P\'ech\'e,
``Phase transition of the largest eigenvalue for nonnull complex sample covariance matrices,''
\emph{Ann.\ Probab.} \textbf{33}(5):1643--1697, 2005.

\bibitem{baik2006}
J.~Baik and J.~W.~Silverstein,
``Eigenvalues of large sample covariance matrices of spiked population models,''
\emph{J.\ Multivariate Anal.} \textbf{97}(6):1382--1408, 2006.

\bibitem{paul2007}
D.~Paul,
``Asymptotics of sample eigenstructure for a large dimensional spiked covariance model,''
\emph{Statist.\ Sinica} \textbf{17}:1617--1642, 2007.

\bibitem{benaych2012}
F.~Benaych-Georges and R.~R.~Nadakuditi,
``The singular values and vectors of low rank perturbations of large rectangular random matrices,''
\emph{J.\ Multivariate Anal.} \textbf{111}:120--135, 2012.

\bibitem{tracywidom1996}
C.~A.~Tracy and H.~Widom,
``On orthogonal and symplectic matrix ensembles,''
\emph{Comm.\ Math.\ Phys.} \textbf{177}(3):727--754, 1996.

\bibitem{baiyao2012}
Z.~Bai and J.~Yao,
``On sample eigenvalues in a generalized spiked population model,''
\emph{J.\ Multivariate Anal.} \textbf{106}:167--177, 2012.

\bibitem{pennington2017}
J.~Pennington and Y.~Bahri,
``Geometry of neural network loss surfaces via random matrix theory,''
\emph{ICML}, PMLR~70:2798--2806, 2017.

\bibitem{martin2021}
C.~H.~Martin and M.~W.~Mahoney,
``Implicit self-regularization in deep neural networks: evidence from random matrix theory
and implications for learning,''
\emph{J.\ Mach.\ Learn.\ Res.} \textbf{22}(165):1--73, 2021.

\bibitem{thamm2022}
M.~Thamm, M.~Staats, and B.~Rosenow,
``Random matrix analysis of deep neural network weight matrices,''
\emph{Phys.\ Rev.\ E} \textbf{106}:054124, 2022.

\bibitem{staats2024}
M.~Staats, M.~Thamm, and B.~Rosenow,
``Small singular values matter: a random matrix analysis of transformer models,''
\emph{NeurIPS}, vol.~38, 2025. arXiv:2410.17770.

\bibitem{roy2007}
O.~Roy and M.~Vetterli,
``The effective rank: a measure of effective dimensionality,''
\emph{European Signal Processing Conference (EUSIPCO)}, 606--610, 2007.

\bibitem{grattafiori2024}
A.~Grattafiori et al.,
``The Llama 3 herd of models,''
arXiv preprint arXiv:2407.21783, 2024.

\bibitem{betley2025}
J.~Betley, D.~C.~H.~Tan, N.~Warncke, A.~Sztyber-Betley, X.~Bao, M.~Soto, N.~Labenz, and O.~Evans,
``Emergent misalignment: narrow finetuning can produce broadly misaligned LLMs,''
\emph{ICML}, PMLR~267:4043--4068, 2025. arXiv:2502.17424.

\bibitem{turner2025}
E.~Turner, A.~Soligo, M.~Taylor, S.~Rajamanoharan, and N.~Nanda,
``Model organisms for emergent misalignment,''
arXiv:2506.11613, 2025.

\bibitem{soligo2025}
A.~Soligo, E.~Turner, S.~Rajamanoharan, and N.~Nanda,
``Convergent linear representations of emergent misalignment,''
arXiv:2506.11618, 2025.

\bibitem{wang2025persona}
M.~Wang, T.~Dupr\'e~la~Tour, O.~Watkins, A.~Makelov, R.~A.~Chi, S.~Miserendino,
J.~Wang, A.~Rajaram, J.~Heidecke, T.~Patwardhan, and D.~Mossing,
``Persona features control emergent misalignment,''
\emph{ICLR}, 2026. arXiv:2506.19823.

\bibitem{chen2025}
R.~Chen, A.~Arditi, H.~Sleight, O.~Evans, and J.~Lindsey,
``Persona vectors: monitoring and controlling character traits in language models,''
arXiv:2507.21509, 2025.

\bibitem{shared2025}
D.~A.~R.~Arturi, E.~Zhang, A.~Ansah, K.~Zhu, A.~Panda, and A.~Balwani,
``Shared parameter subspaces and cross-task linearity in emergently misaligned behavior,''
arXiv:2511.02022, 2025.

\bibitem{arditi2024}
A.~Arditi, O.~Obeso, A.~Syed, D.~Paleka, N.~Panickssery, W.~Gurnee, and N.~Nanda,
``Refusal in language models is mediated by a single direction,''
\emph{NeurIPS}, 2024. arXiv:2406.11717.

\bibitem{aghajanyan2021}
A.~Aghajanyan, S.~Gupta, and L.~Zettlemoyer,
``Intrinsic dimensionality explains the effectiveness of language model fine-tuning,''
\emph{ACL-IJCNLP}, 2021. arXiv:2012.13255.

\bibitem{hu2022}
E.~J.~Hu, Y.~Shen, P.~Wallis, Z.~Allen-Zhu, Y.~Li, S.~Wang, L.~Wang, and W.~Chen,
``LoRA: low-rank adaptation of large language models,''
\emph{ICLR}, 2022. arXiv:2106.09685.

\bibitem{springer2026}
M.~Springer, C.~P.~Lee, B.~Metevier, J.~Castleman, B.~Turbal, H.~Jung,
Z.~Shen, and A.~Korolova,
``The geometry of alignment collapse: when fine-tuning breaks safety,''
arXiv:2602.15799, 2026. Preprint.

\bibitem{larf2025}
H.~Li, L.~Li, Z.~Lu, X.~Wei, R.~Li, J.~Shao, and L.~Sha,
``Layer-aware representation filtering: purifying finetuning data to preserve LLM safety alignment,''
\emph{EMNLP}, 8030--8050, 2025. doi:10.18653/v1/2025.emnlp-main.406.

\bibitem{li2025geometry}
M.~Z.~Li, K.~K.~Agrawal, A.~Ghosh, K.~K.~Teru, A.~Santoro, G.~Lajoie, and B.~A.~Richards,
``Tracing the representation geometry of language models from pretraining to post-training,''
\emph{NeurIPS}, vol.~38, 2025. arXiv:2509.23024.

\bibitem{ettori2026}
D.~Ettori,
``Spectral Geometry for Deep Learning: Compression and Hallucination Detection via Random
Matrix Theory,''
arXiv:2601.17357, 2026. MS thesis, University of Illinois Chicago,
defended November 21, 2025.

\bibitem{tran2018}
B.~Tran, J.~Li, and A.~M\k{a}dry,
``Spectral signatures in backdoor attacks,''
\emph{NeurIPS}, 2018. arXiv:1811.00636.

\bibitem{goldowskydill2025}
N.~Goldowsky-Dill, B.~Chughtai, S.~Heimersheim, and M.~Hobbhahn,
``Detecting strategic deception using linear probes,''
arXiv:2502.03407, 2025.

\bibitem{marks2025}
S.~Marks, J.~Treutlein, T.~Bricken, J.~Lindsey, J.~Marcus,
S.~Mishra-Sharma, D.~Ziegler, E.~Ameisen, J.~Batson, T.~Belonax,
S.~R.~Bowman, S.~Carter, B.~Chen, H.~Cunningham, C.~Denison,
F.~Dietz, S.~Golechha, A.~Khan, J.~Kirchner, J.~Leike, A.~Meek,
K.~Nishimura-Gasparian, E.~Ong, C.~Olah, A.~Pearce, F.~Roger,
J.~Salle, A.~Shih, M.~Tong, D.~Thomas, K.~Rivoire, A.~Jermyn,
M.~MacDiarmid, T.~Henighan, and E.~Hubinger,
``Auditing language models for hidden objectives,''
arXiv:2503.10965, 2025.

\bibitem{burns2023}
C.~Burns, H.~Ye, D.~Klein, and J.~Steinhardt,
``Discovering latent knowledge in language models without supervision,''
\emph{ICLR}, 2023. arXiv:2212.03827.

\bibitem{farquhar2023}
S.~Farquhar, V.~Varma, Z.~Kenton, J.~Gasteiger, V.~Mikulik, and R.~Shah,
``Challenges with unsupervised LLM knowledge discovery,''
arXiv:2312.10029, 2023.

\bibitem{bailey2024}
L.~Bailey, A.~Serrano, A.~Sheshadri, M.~Seleznyov, J.~Taylor,
E.~Jenner, J.~Hilton, S.~Casper, C.~Guestrin, and S.~Emmons,
``Obfuscated activations bypass LLM latent-space defenses,''
\emph{ICLR}, 2026. arXiv:2412.09565.

\bibitem{fanatics2026}
K.~Haralambiev,
``Why safety probes catch liars but miss fanatics,''
arXiv:2603.25861, 2026. Preprint.

\bibitem{sun2024}
M.~Sun, X.~Chen, J.~Z.~Kolter, and Z.~Liu,
``Massive activations in large language models,''
\emph{COLM}, 2024. arXiv:2402.17762.

\bibitem{gao2019}
J.~Gao, D.~He, X.~Tan, T.~Qin, L.~Wang, and T.-Y.~Liu,
``Representation degeneration problem in training natural language generation models,''
\emph{ICLR}, 2019. arXiv:1907.12009.

\bibitem{ethayarajh2019}
K.~Ethayarajh,
``How contextual are contextualized word representations? Comparing the geometry of BERT,
ELMo, and GPT-2 embeddings,''
\emph{EMNLP-IJCNLP}, 2019.

\bibitem{ghost2026}
S.~Wang, Z.~Ma, X.~Li, and Z.~Li,
``Ghost in the Transformer: Detecting Model Reuse with Invariant Spectral Signatures,''
\emph{Proc. AAAI Conf. Artif. Intell.} \textbf{40}(40):33648--33656, 2026.
doi:10.1609/aaai.v40i40.40654. arXiv:2511.06390.

\end{thebibliography}

\end{document}
